Potential competition in alpine bumble bees is mitigated by imperfect trait-matching and subpartitioning within floral morphotypes

### 2021-03-08 ms_2/ms_2_v2.Rmd
improved niche overlap analysis

All else held equal, species coexistence is favored when intraspecific competition exceeds interspecific competition, and this stabilizing inequality is favored to the extent that species differ in dietary niche [@Abrams1977-zp; @Chesson2000-aw]. Bumble bees are exceptionally versatile foragers whose ability to solve the puzzles posed by complex flower morphologies [Plowright1984-lc], or indeed by human investigators [@Alem2016-rn; @Loukola2017-bt], renders almost any flower potentially accessible. Nevertheless, the flexibility of bumble bee foraging is tuned to floral reward, which depends on both the intrinsic reward of a given flower and the costs associated with exploiting it. These costs are closely related to the morphological fittedness between bumble bees and their floral hosts, which consists principally in the compatibility between bumble bee tongue length and floral corolla depth. Bumble bee species can differ markedly in tongue length, and floral handling time increases both for short-tongued bumble bees foraging on deep flowers and long-tongued bees foraging on shallow flowers [@Inouye1980-mv], which leads to an association between tongue length and floral niche in bumble bee communities [e.g. @Inouye1978-qi; @Inouye1980-mv]. This association between tongue length and floral morphology, though, is imperfect [e.g. @Ranta1980-ok; @Goulson2008-oe], and access to floral rewards in some flowers involves learning to manipulate floral structures that are challenging irrespective of tongue length [@Laverty1994-hm]. For example, the characteristic Papilionaceous floral morphology of legumes requires a bee to separate tightly appressed keel petals to access anthers and nectaries [@Westerkamp1997-hs], and the narrowly campanulate flowers of many Ericaceous flora pose a significant barrier to large-bodied visitors like bumble bees [@Wallace1977-kz]. Thus the dietary niches of bumble bee species in a given ecological context are emergent products of innate floral biases, rewards experienced during foraging, and learned flower handling skills, all of which contribute to the phenomenon of interspecific niche overlap upon which interspecific competition is predicated.

@@

Rather than considering various floral morphological traits independently, we rely on the synthesis of @Kugler1970-gb, who developed a typology consisting of ten primary morphotypes and nested subdivision of each. Kugler’s typology was informed by laboratory studies of bumble bee flower choice [@Kugler1943-uy], and there is a precedent for analyzing bumble bee visitation in terms of Kugler’s morphotypes [@Neumayer1999-il; @Schneller2014-ry].

Kugler morphotype classifications of the plants in our study system were accessed via the BIOFLOR database (Klotz, Kühn, and Durka 2002). Morphotypes were simplified to the primary classes numbered 0-10: 0 = not applicable (includes wind-pollinated flora such as grasses), 1 = disc- and bowl-shaped flowers, 2 = funnel flowers, 3 = bell-shaped flowers, 4 = stalk-disc flowers, 5 = lip flowers, 6 = flag flowers, 7 = flower heads, 8 = spike flowers, 9 = brush flowers, 10 = trap flowers **(Figure 1)**. In a few cases, a plant species in our data set was not found in the BIOFLOR database. In most cases, morphotypes are constant at the genus level, and missing morphotypes were added based on the morphotypes found for congeners. The genus *Gentiana* consisted of both group 2 and group 4 representatives, so the missing species *Gentiana aspera* and *Gentiana ciliata* were classified based on visual comparison with their classified congeners. In the case of the genus *Phyteuma*, we followed Neumayer and Paulus (1999) and classified it as a group 7 flower head. Morphotypes 9 and 10 were extremely rare in our study system and were therefore omitted from all analyses. Plants belonging to morphotype 0 were typically not recorded during floral surveys due to their lack of conspicuous floral displays, so this morphotype was also omitted from all analyses.

@@

Prior to analysis, we aligned each visitation sample with a corresponding floral survey sample. In most cases, floral surveying was conducted jointly with bumble bee observation for each site and date. Occasionally, though, floral surveying could not be completed on the same day as bumble bee observation, resulting in a small temporal offset between visitation samples and floral survey samples. In these cases, visitation samples were paired with the nearest available floral survey data within a maximum distance of 7 days. Visitation samples lacking corresponding floral survey data within 7 days were omitted from further analysis. In a few cases, two consecutive visitation samples at a given site shared the same nearest floral survey sample; in these cases, the visitation samples were first pooled, then aligned to their shared nearest floral survey sample. In rare cases, a given floral species was recorded during visitation sampling but missed during floral surveying. These species were retroactively assigned the minimum floral area value of 0.01 m^2.

Visitation data were extremely right-skewed because the majority of available flora received zero visits during a given sampling period. To facilitate model fitting, we created a binary version of visitation rate in which visited flora were assigned to 1 and unvisited flora to 0. We then modeled binary visitation using generalized models with binomial or Bernoulli distribution families. The conditional effects inferred from these models should, therefore, be interpreted as effects on visitation probability rather than visitation frequency.

While our study spanned three years, we chose not to model year effects. Differences in sampling intensity and duration across years, for the reasons described above, would make comparisons across years difficult to interpret, and we deemed these comparisons not worth the cost of added model complexity. This decision entails the caveats that our analysis cannot detect year-to-year variation in floral biases and niche overlap, and main effects cannot be distinguished from annual anomalies. With only three years of data, though, such annual inferences would be of little value.

<!-- State more explicityl how the models should be interpreted withr expect to bias -->

@@

@@

### Model 5: Niche overlap by level

### Model 6: Genus-level overlap by species-pair

### Model 7: Community-mean genus-level niche overlap by elevation

We analyze both pairwise niche overlap and community mean niche overlap, since low pairwise niche overlap could conceal a high potential for diffuse competition [@Pianka1974-qk; MacArthur1972-qz]

### Model 8: Pairwise genus-level niche overlap by elevation

@@

All analyses were conducted in `R` `r citep(citation())`. Bayesian multilevel models were implemented in Stan (Stan Development Team 2021) via the package `brms` `r citep(citation("brms"))`, and visualizations were constructed with the packages `tidybayes` `r citep(citation("tidybayes"))` and `ggdist` `r citep(citation("ggdist"))`. Ordination analysis and permutation testing were performed with the package `vegan` `r citep(citation("vegan"))` and visualized using the packages `ggplot2` `vegan` `r citep(citation("ggplot2"))` and `ggvegan` (Simpson 2019). GAMs were implemented with the package `mgcv` [@Wood_2017] and visualized with the package `mgcViz` [@Fasiolo2018aa]. Data handling and visualization was performed with the `tidyverse` suite of R packages `r citep(citation("tidyverse"))`. Annotated R code is available in the Supplementary Material.

@@

```{r fig5, echo = FALSE,  fig.cap = paste(" Principal coordinates ordination of bumble bee visitation pooled across all samples.")}
knitr::include_graphics("../output/pcoa_aggregate.png")
```

@@

>>>>

All else held equal, species coexistence is favored when intraspecific competition exceeds interspecific competition, and this stabilizing inequality is favored to the extent that species differ in dietary niche [@Abrams1977-zp; @Chesson2000-aw]. Bumble bees are exceptionally versatile foragers whose ability to solve the puzzles posed by complex flower morphologies [Plowright1984-lc], or indeed by human investigators [@Alem2016-rn; @Loukola2017-bt], renders almost any flower potentially accessible. Nevertheless, the flexibility of bumble bee foraging is tuned to floral reward, which depends on both the intrinsic reward of a given flower and the costs associated with exploiting it. These costs are closely related to the morphological fittedness between bumble bees and their floral hosts, which consists principally in the compatibility between bumble bee tongue length and floral corolla depth. Bumble bee species can differ markedly in tongue length, and floral handling time increases both for short-tongued bumble bees foraging on deep flowers and long-tongued bees foraging on shallow flowers [@Inouye1980-mv], which leads to an association between tongue length and floral niche in bumble bee communities [e.g. @Inouye1978-qi; @Inouye1980-mv]. This association between tongue length and floral morphology, though, is imperfect [e.g. @Ranta1980-ok; @Goulson2008-oe], perhaps in part because access to floral rewards in some flowers involves learning to manipulate floral structures that are challenging irrespective of tongue length [@Laverty1994-hm]. For example, the characteristic Papilionaceous floral morphology of legumes requires a bee to separate tightly appressed keel petals to access anthers and nectaries [@Westerkamp1997-hs], and the narrowly campanulate flowers of many Ericaceous flora pose a significant barrier to large-bodied visitors like bumble bees [@Wallace1977-kz]. Thus the dietary niches of bumble bee species in a given ecological context are emergent products of innate floral biases, rewards experienced during foraging, and learned flower handling skills, all of which contribute to the phenomenon of interspecific niche overlap upon which interspecific competition is predicated.

@@

Rather than considering various floral morphological traits independently, we relied on the synthesis of @Kugler1970-gb, who developed a typology consisting of ten primary morphotypes and nested subdivision of each. Kugler’s typology was informed by laboratory studies of bumble bee flower choice [@Kugler1943-uy], and there is a precedent for analyzing bumble bee visitation in terms of Kugler’s morphotypes [@Neumayer1999-il; @Schneller2014-ry].

Kugler morphotype classifications of the plants in our study system were accessed via the BIOFLOR database (Klotz, Kühn, and Durka 2002). Morphotypes were simplified to the primary classes numbered 0-10: 0 = not applicable (includes wind-pollinated flora such as grasses), 1 = disc- and bowl-shaped flowers, 2 = funnel flowers, 3 = bell-shaped flowers, 4 = stalk-disc flowers, 5 = lip flowers, 6 = flag flowers, 7 = flower heads, 8 = spike flowers, 9 = brush flowers, 10 = trap flowers **(Figure 1)**. In a few cases, a plant species in our data set was not found in the BIOFLOR database. In most cases, morphotypes are constant at the genus level, and missing morphotypes were added based on the morphotypes found for congeners. The genus *Gentiana* consisted of both group 2 and group 4 representatives, so the missing species *Gentiana aspera* and *Gentiana ciliata* were classified based on visual comparison with their classified congeners. In the case of the genus *Phyteuma*, we followed Neumayer and Paulus (1999) and classified it as a group 7 flower head. Morphotypes 9 and 10 were extremely rare in our study system and were therefore omitted from all analyses. Plants belonging to morphotype 0 were typically not recorded during floral surveys due to their lack of conspicuous floral displays, and they were rarely visited by bumble bees, so this morphotype was also omitted from all analyses.

@@

Prior to analysis, we aligned each visitation sample with a corresponding floral survey sample. In most cases, floral surveying was conducted jointly with bumble bee observation for each site and date. Occasionally, though, floral surveying could not be completed on the same day as bumble bee observation, resulting in a small temporal offset between visitation samples and floral survey samples. In these cases, visitation samples were paired with the nearest available floral survey data within a maximum distance of 7 days. Visitation samples lacking corresponding floral survey data within 7 days were omitted from further analysis. In a few cases, two consecutive visitation samples at a given site shared the same nearest floral survey sample; in these cases, the visitation samples were first pooled, then aligned to their shared nearest floral survey sample. In rare cases, a given floral species was recorded during visitation sampling but missed during floral surveying. These species were retroactively assigned the minimum floral area value of 0.01 m^2^.

Visitation data were extremely right-skewed because the majority of available flora received zero visits during any given sampling period. To facilitate model fitting, we created a binary version of visitation rate in which visited flora were assigned to 1 and unvisited flora to 0. We then modeled binary visitation using generalized models with binomial or Bernoulli distribution families. The conditional effects inferred from these models must, therefore, be interpreted as effects on visitation probability rather than visitation frequency.

We calculated niche overlap with the goal of focusing on elevational patterns, sacrificing temporal resolution to pool observations within site. First, we calculated each bumble bee species' proportional visitation frequency at the nested levels of floral morphotype, plant family, and plant genus. Then, we calculated, for each bumble bee species, the mean proportional visitation rate to each morphotype/family/genus across all sampling dates within each site. Using these species-level visitation matrices, we calculated pairwise niche overlap as 1 - Horn-Morisita distance (calculated separately for morphotype-, family-, and genus-level matrices) for all species pairs within each site. This approach was chosen for its normalizing effect on the highly variable "snapshot" visitation data generated by our individual transect walks, but it has the disadvantage of conflating niche overlap due to similar floral preference and niche overlap due to similar temporal (i.e. seasonal) activity patterns.

<!-- Maybe explain why we only went down to genus instead of species -->

While our study spanned three years, we chose not to model year effects. Differences in sampling intensity and duration across years, for the reasons described above, would make comparisons across years difficult to interpret. This decision entails the caveats that our analysis cannot detect year-to-year variation in floral biases and niche overlap, and main effects cannot be distinguished from annual anomalies. With only three years of data, though, such annual inferences would be of little value, and we deemed these comparisons not worth the cost of added model complexity.

<!-- State more explicityl how the models should be interpreted with respect to bias -->

@@

<!-- Something about priors -->

@@

### Model 5: Niche overlap by level and elevation

We modeled niche overlap as a function of level (morphotype, family, genus) and elevation using hierarchical generalized additive models (HGAMs). First, we calculated community mean niche overlap by taking the mean niche overlap (within level) across all species pairs for each site.  Level was included as both a fixed intercept effect and a smoothing factor in conjunction with elevation, following the type "I" model form described by @Pedersen2019-kf. To account for the proportional nature of our data, we used a beta distribution family with a logit link function.

### Model 6: Genus-level niche overlap by tongue length and elevation

To focus on genus-level niche overlap as a function of tongue length class, we generated a binary tongue-difference variable for each species pair: "same" for species belonging to the same tongue length class and "different" for species belonging to different tongue length classes. We then took the mean genus-level niche overlap across all species pairs within tongue-difference class and site. Then, we modeled genus-level niche overlap using tongue-difference as both a fixed intercept effect and a smoothing factor in conjunction with elevation, again following the type "I" model form @Pedersen2019-kf.

@@

All analyses were conducted in `R` `r citep(citation())`. Bayesian multilevel models were implemented in Stan (Stan Development Team 2021) via the package `brms` `r citep(citation("brms"))`, and visualizations were constructed with the packages `tidybayes` `r citep(citation("tidybayes"))` and `ggdist` `r citep(citation("ggdist"))`. Ordination analysis and permutation testing were performed with the package `vegan` `r citep(citation("vegan"))` and visualized using the packages `ggplot2` `vegan` `r citep(citation("ggplot2"))` and `ggvegan` (Simpson 2019). HGAMs were implemented with the package `mgcv` [@Wood_2017] and visualized with the package `mgcViz` [@Fasiolo2018aa]. Data handling and visualization was performed with the `tidyverse` suite of R packages `r citep(citation("tidyverse"))`. Annotated R code is available in the Supplementary Material.

@@

```{r fig5, echo = FALSE,  fig.cap = paste("Principal coordinates ordination of bumble bee visitation pooled across all samples.")}
knitr::include_graphics("../output/pcoa_aggregate.png")
```

## Niche overlap

### Niche overlap through elevation by level

Morphotype-level niche overlap was nearly double genus-level niche overlap throughout the whole elevation gradient, with family-level niche overlap taking intermediate values. The fixed intercepts of morphotype- and genus-level niche overlap each differed significantly (in opposite directions) from the reference level of family-level niche overlap (p << 0.001). The conditional smooths of each level were significantly different from flat lines (p < 0.001) and generally mirrored one another in shape **(Figure 8)**. Niche overlap was highest at the low end of the elevation gradient, decreased steeply from 600-1200 m, and leveled off or declined weakly for the remainder of the gradient, though morphotype-level niche overlap exhibited a second peak just above 1500 m that was not evident in genus- or family-level niche overlap. The model explained 78.4% of total deviance (adjusted R^2^ = 0.743).

```{r fig8, echo = FALSE,  fig.cap = paste("Niche overlap by level through elevation. Points show empirical niche overlap values, lines show GAM smooths, and shaded bands show the standard errors of the mean")}
knitr::include_graphics("../output/niche_gam00_plot1.png")
```

### Genus-level niche overlap by tongue length and elevation

Whether a species pair belonged to the same or different tongue length classes had no effect on genus-level niche overlap, either in terms of intercept (p = 0.18) or response to elevation **(Figure 9)**. The model explained 36.5% of total deviance (adjusted R^2^ = 0.316).

```{r fig9, echo = FALSE,  fig.cap = paste("Niche overlap by tongue length class difference through elevation. Lines show GAM smooths and shaded bands show standard errors of the mean")}
knitr::include_graphics("../output/niche_gam20_plot1.png")
```

@@

We analyze community mean niche overlap as an indication of diffuse competition [@Pianka1974-qk; MacArthur1972-qz]
